Automatic Repulsive Clustering \\for Finding Separable Linear Subspace

### 2016-11-28 docs/draft.tex
updated draft

\usepackage[utf8]{inputenc}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

@@

To ensure identifiability, we regularize $V_{i,j}\sim N(0,1)$. It is possible to  maximize the log-likelihood alternatively over $\V$ and $\X$, however, this would underestimate the variability of the $\V$, leading to suboptimal result. Instead, we treat $\V$ as latent variable and use EM algorithm for optimization.

Note the conditional distribution for $\V$ is:
\be
\V_{.,j} \stackrel{indep}{\sim} N\left( (\X'\X+\I)^{-1}\X'\Y_{.,j}, \sigma^2 (\X'\X+\I)^{-1}    \right)
\ee
for $j=1\ldots p$.

This leads to computing the expectation:
\be
\E\V =  (\X'\X+\I)^{-1}\X'\Y,  \qquad\E\V \V' = p\sigma^2 (\X'\X+\I)^{-1} + (\X'\X+\I)^{-1}\X'\Y \Y'\X (\X'\X+\I)^{-1}
\ee

@@

@@

Then the whole estimation can be carried out by alternating in the two steps.

\section{Simulation}

Data generation:

@@

\begin{figure}[H]
 % \centering
  \begin{subfigure}[b]{0.49\textwidth}
 \includegraphics[width=0.8\textwidth]{pics/originalY2d.pdf}
  \caption{The 3 component view in the first 2 dimensions.}
\end{subfigure}
  \hfill
   \begin{subfigure}[b]{0.49\textwidth}
 \includegraphics[width=0.8\textwidth]{pics/ld_arc_truth.pdf}
  \caption{The true latent 2d separable subspace view.}
\end{subfigure}

 \begin{subfigure}[b]{0.49\textwidth}
 \includegraphics[width=0.8\textwidth]{pics/pca_gmm.pdf}
  \caption{GMM on principle components}
\end{subfigure}
  \hfill
   \begin{subfigure}[b]{0.49\textwidth}
 \includegraphics[width=0.8\textwidth]{pics/pca_arc.pdf}
  \caption{ARC on principle components}
\end{subfigure}

 \begin{subfigure}[b]{0.49\textwidth}
 \includegraphics[width=0.8\textwidth]{pics/ld_gmm.pdf}
  \caption{GMM on flexible reduced Dimension}
\end{subfigure}
  \hfill
   \begin{subfigure}[b]{0.49\textwidth}
 \includegraphics[width=0.8\textwidth]{pics/ld_arc.pdf}
  \caption{ARC on flexible reduced Dimension}
\end{subfigure}

 \end{figure}

\section{Application}
\section{Discussion}

>>>>

\usepackage[utf8]{inputenc}
%\usepackage{algorithm}
%\usepackage[noend]{algpseudocode}
\usepackage[]{algorithm2e}

@@

To ensure identifiability, we regularize $V_{i,j}\sim N(0,\nu)$, where $\nu$ is a large variance (e.g. $10^6$). It is possible to  maximize the log-likelihood alternatively over $\V$ and $\X$, however, this would underestimate the variability of the $\V$, leading to suboptimal result. Instead, we treat $\V$ as latent variable and use EM algorithm for optimization.

Note the conditional distribution for $\V$ is:
\be
\V_{.,j} \stackrel{indep}{\sim} N\left( (\X'\X+\I \nu)^{-1}\X'\Y_{.,j}, \sigma^2 (\X'\X+\I\nu)^{-1}    \right)
\ee
for $j=1\ldots p$.

This leads to computing the expectation:
\be
\E\V =  (\X'\X+\I \nu)^{-1}\X'\Y,  \qquad\E\V \V' = p\sigma^2 (\X'\X+\I \nu)^{-1} + (\X'\X+\I \nu)^{-1}\X'\Y \Y'\X (\X'\X+\I \nu)^{-1}
\ee

@@

As the loss function, the expected log-likelihood is:

\be
\bb E \log L = &  - \frac{1}{2}  [    \{\mbox{vec} (\Y)'\mbox{vec} (\Y)  -2 \mbox{vec}  ( \hat\X')' \mbox{vec}( \E\V \Y') + \mbox{vec}  ( \hat\X')' \mbox{vec}( \E\V\V' {\hat \X}' )\}/{\sigma^2} +\\
& {\sum_i^n (\X_{i,.} - \Z_{i,.} \bmu)' \Sigma^{-1}  (\X_{i,.} - \Z_{i,.} \bmu)} + n\log \det  \bSigma + np \log \sigma^2 ]
\ee

@@

Then the whole estimation can be carried out by alternating in the two steps. To summarize, the estimating algorithm is shown in Algorithm~\ref{arc_algo}.

\begin{algorithm}[H]
 initialization\;
 \While{Expected likelihood $\bb E \log L$ change $>$ threshold}{
  Using $\hat \Z \hat \bmu$, compute expectation: $\bb E \V$ and $\bb E \V\V'$\;
  Using expected values, maximize: $\hat \X$ \;
   \While{change in $\hat \bmu$ $>$ threshold}{
   		  Compute MLEs: $\hat \bmu$ and $\hat \bSigma$\;
		  Compute Most Probable Assignment: $\hat \Z$ \;
	}
 Compute expected likelihood $\bb E \log L$\;
 }
 \caption{Estimation Algorithm for Automatic Repulsive Clustering}
 \label{arc_algo}
\end{algorithm}

\section{Experiments}

\subsection{Synthetic Data}
Data generation:

@@

\begin{figure}[H]
 % \centering
  \begin{subfigure}[b]{0.32\textwidth}
 \includegraphics[width=0.8\textwidth]{pics/simX}
  \caption{The lower dimensions used in data generation.}
\end{subfigure}
  \hfill
   \begin{subfigure}[b]{0.32\textwidth}
 \includegraphics[width=0.8\textwidth]{pics/simPCAX}
  \caption{The lower dimensions discovered by PCA.}
\end{subfigure}
  \hfill
 \begin{subfigure}[b]{0.32\textwidth}
 \includegraphics[width=0.8\textwidth]{pics/simARCX}
  \caption{The lower dimensions discovered by ARC.}
\end{subfigure}
 \end{figure}

\subsection{Real Data}

MNIST:

Four digits: $0,3,7,9$. Pick $100$ samples for each digit.

Compute average ARI over $10$ tests.

\begin{table}[H]
\centering
	\begin{tabular}{| l | r |}
	\hline
			Model & $ARI$ \\
	\hline
			GMM & 0.50 \\
			PCA + GMM &  0.38  \\
			ARC		& {\bf 0.57} \\
			\hline
	\end{tabular}
\end{table}

\section{Discussion}
